k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2023-05-08 paper.Rmd
still editing for revision 1

[^dissimilarityfromfeatures]: It is still possible to convert the attribute data into a dissimilarity matrix, e.g. by computing all pairwise Euclidean distances, and then optimize a distance-based anticlustering criterion to obtain a partitioning. For Likert scale data or binary attributes, the Manhattan distance may be preferred over the Euclidean distance; for mixed-type attributes, it is even possible to employ a combined distance metric [@rokach2005].

@@

Note that $\mathit{SSE_{Var}}$ has the same form as the standard $\mathit{SSE}$. However, the input data reflects the squared difference between each data point and the attribute mean instead of the raw data points. Maximizing $\mathit{SSE_{Var}}$ can be used to obtain groups that have similar variances across the $P$ attributes. As a caveat, however, it is crucial to note that maximizing $\mathit{SSE_{Var}}$ only reliably yields similar group variances if the group centroids $\overline{\mathbf{x}}^{(k)}$ ($k = 1, \dots, K$) are close to the overall data centroid $\overline{\mathbf{x}}$. This is because the within-group variance is calculated on the basis of the group centroid $\overline{\mathbf{x}}^{(k)}$, but the new variables $y_{ij}$ are calculated using the overall centroid $\overline{\mathbf{x}}$ as reference point. To ensure that $\mathit{SSE_{Var}}$ is a valid criterion, we must therefore ensure that the group centroids of the original variables are close to the overall data centroid (and therefore to each other). Ensuring that the group means are close to the overall mean is accomplished by maximizing the classical $\mathit{SSE}$, i.e., via standard k-means anticlustering [@spath1986]. Combining the maximization of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ at the same time leads to the k-plus approach for anticlustering, which is detailed in the next section.

@@

By maximizing $\mathit{SSE_{kplus}}$ we seek to obtain a partitioning where means as well as variances are similar between groups, while the weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of means and variances. However, research on multiobjective optimization indicates that solutions oftentimes fail to satisfactorily address several criteria, if the underlying objectives are in conflict rather than in alignment [@ferligoj1992direct]. As $\mathit{SSE_{kplus}}$ is a novel criterion, it is so far unknown if the two constituting criteria are aligned. That is, it has yet to be demonstrated that maximizing $\mathit{SSE_{kplus}}$ can yield a partition where means as well as variances are similar between groups. A motivating example illustrates that aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatter plot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. The criteria seem to be rather well aligned: There is a high concentration of partitions where both the $\mathit{SSE}$ and the $\mathit{SSE_{Var}}$ have high values. At the same time, there are many partitions that fulfil only one of the criteria, but fail to simultaneously fulfil the other criterion. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (represented by the triangular symbol) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes the unweighted sum of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$, which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, even using a very simple unweighted sum approach, the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ can be fulfilled to an astonishing degree.

@@

To answer the question if different anticlustering partitions are similar or dissimilar, Table 3 illustrates the degree of overlap between anticlustering partitions that were identified in Simulation 2. Overall, there was hardly any similarity between anticlustering partitions that were obtained via optimization of the different criteria. The average ARI was almost 0 in all cases---indicating no similarity of partitions whatsoever. Even the pairwise ARI of the two k-plus criteria was fractional, although the criteria overlapped considerably in their optimization goal. This constitutes an interesting result, illustrating that high between-group similarity can be achieved via very different partitions.

@@

As an additional outlook, it should be noted that because the covariance is defined as an expected value (i.e., an average), the k-plus logic can be extended to minimize differences with regard to covariances, a possibility that was not investigated in the current paper. Similar to addressing distribution moments, covariances could be incorporated by appending additional variables to a data set. In this case, it would be necessary to compute an additional variable for each of pair of variables, because the covariance is the product of the differences between two variables and their respective means.

>>>>

[^dissimilarityfromfeatures]: It is still possible to convert the attribute data into a dissimilarity matrix, e.g. by computing all pairwise Euclidean distances, and then optimize a distance-based anticlustering criterion to obtain a partitioning. For Likert scale data or binary attributes, the Manhattan distance may be preferred over the Euclidean distance; for mixed-type attributes, it is even possible to compute a combined distance metric [@rokach2005].

@@

Note that $\mathit{SSE_{Var}}$ has the same form as the standard $\mathit{SSE}$. However, the input data reflects the squared difference between each data point and the attribute mean instead of the raw data points. Maximizing $\mathit{SSE_{Var}}$ can be used to obtain groups that have similar variances across the $P$ attributes. As a caveat, however, it is crucial to note that maximizing $\mathit{SSE_{Var}}$ only reliably yields similar group variances if the group centroids $\overline{\mathbf{x}}^{(k)}$ ($k = 1, \dots, K$) are close to the overall data centroid $\overline{\mathbf{x}}$. This is because the within-group variance is calculated on the basis of the group centroid $\overline{\mathbf{x}}^{(k)}$, but the new variables $y_{ij}$ are calculated using the overall centroid $\overline{\mathbf{x}}$ as reference point. Using $\overline{\mathbf{x}}$ as reference point is however necessary because the group means $\overline{\mathbf{x}}^{(k)}$ are determined by the composition of the groups $(C_1, \ldots, C_K)$, which is the end result of the optimization and not known in advance. To ensure that $\mathit{SSE_{Var}}$ is a valid criterion, we must therefore ensure that the group centroids of the original variables are close to the overall data centroid (and therefore to each other). Ensuring that the group means are close to the overall mean is accomplished by maximizing the classical $\mathit{SSE}$, i.e., via standard k-means anticlustering (Späth, 1986). Combining the maximization of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ at the same time leads to the k-plus approach for anticlustering, which is detailed in the next section.

@@

By maximizing $\mathit{SSE_{kplus}}$, we seek to obtain a partitioning where means as well as variances are similar between groups, while the weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of means and variances. However, research on multiobjective optimization indicates that solutions oftentimes fail to satisfactorily address several criteria if the underlying objectives are in conflict rather than in alignment [@ferligoj1992direct]. As $\mathit{SSE_{kplus}}$ is a novel criterion, it is so far unknown if the two constituting criteria are aligned. That is, it has yet to be demonstrated that maximizing $\mathit{SSE_{kplus}}$ can yield a partition where means as well as variances are similar between groups. A motivating example illustrates that aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatter plot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. The criteria seem to be rather well aligned: There is a high concentration of partitions where both the $\mathit{SSE}$ and the $\mathit{SSE_{Var}}$ have high values. At the same time, there are many partitions that fulfil only one of the criteria, but fail to simultaneously fulfil the other criterion. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (represented by the triangular symbol) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes the unweighted sum of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$, which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, even using a very simple unweighted sum approach, the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ can be fulfilled to an astonishing degree.

@@

To answer the question if different anticlustering partitions are similar or dissimilar, Table 3 illustrates the degree of similarity between anticlustering partitions that were identified in Simulation 2. Overall, there was hardly any similarity between anticlustering partitions that were obtained via optimization of the different criteria. The average ARI was almost 0 in all cases---indicating no similarity of partitions whatsoever. Even the pairwise ARI of the two k-plus criteria was fractional, although the criteria overlapped considerably in their optimization goal. This constitutes an interesting result, illustrating that high between-group similarity can be achieved via very different partitions.

@@

As an additional outlook, it should be noted that because the covariance is defined as an expected value (i.e., an average), the k-plus logic can be extended to minimize differences with regard to covariances. Similar to addressing distribution moments, covariances could be incorporated by appending additional variables to a data set. In this case, it would be necessary to compute an additional variable for each of pair of variables, because the covariance is the product of the differences between two variables and their respective means. Future research should investigate how covariances can properly be included with k-plus anticlustering.
